---
title: Code Output
format:
  closeread-html:
    code-tools: true
---

Closeread sections can showcase a broad range of code output as stickies.

1. Source code and code output (this works)
1. Code output (this works)
1. Source code and plot output (this doesn't work)

:::{.cr-section}

This sticky shows both the source code and the code output. @cr-codein-codeout

You can add highlighting to the source code but not the output (this may be added in a future release). [@cr-codein-codeout]{highlight="3,4"}

:::{#cr-codein-codeout}
```{r}
#| echo: true
#| message: false
library(dplyr)
library(palmerpenguins)
penguins |>
  count(species, island)
```
:::

This sticky shows only the code output. @cr-codeout

:::{#cr-codeout}
```{r}
#| echo: false
#| message: false
library(dplyr)
library(palmerpenguins)
penguins |>
  count(species, island)
```
:::


This sticky shows the source code for a plot as well as the plot. @cr-codein-plotout

:::{#cr-codein-plotout}
```{r}
#| echo: true
#| message: false
#| warning: false
# This sticky blows up the grid layout
library(ggplot2)
ggplot(penguins, aes(x = bill_length_mm,
                     y = bill_depth_mm)) +
  geom_point()
```
:::

:::